---
title: "ACNC_2025_data_cleaning_summary"
author: "Cynthia Perez"
date: "2025-08-25"
output: html_document
---

## Create a copy of these data, then create a smaller data set that only includes those Glowing samples from 6mo and from 24mo [G-XXX & G-XXX 24month].
```{r data cleaning}
library(readxl)
library(tidyverse)
library(ggpubr)
ACNC <- read_excel("Final Results ACNC 2025.xlsx", sheet = 1) 
thresholds <- read_excel("Final Results ACNC 2025.xlsx", sheet = 2) # I created this sheet and transposed it on excel 

#replacing space from header names with a _
names(ACNC) <- gsub(" ", "_", names(ACNC))
thresholds$PFAS <- gsub(" ", "_", thresholds$PFAS)
thresholds = data.frame(thresholds)
rownames(thresholds) = thresholds$PFAS

#extracting sample names with G- and 24 mo in name
ACNC_24mo = ACNC %>% filter(str_detect(Sample_Name, "G-") & str_detect(Sample_Name, " 24mo") ) 
dim(ACNC_24mo)

#dropping 24 mo from name to filter for 6 mo 
IDs = str_extract(ACNC_24mo$Sample_Name, "^[^ ]+")

ACNC_6mo = ACNC %>% filter(Sample_Name %in% IDs)
dim(ACNC_6mo)
```

# removing duplicates and not shared IDs 
```{r}
# identifying duplicates in full data set
ACNC$Sample_Name[duplicated(ACNC$Sample_Name)]

#samples that do not have a 24 mo but no 6 mo
remove = IDs[!(IDs %in% ACNC_6mo$Sample_Name)]
remove
#all 6 mo samples have a 24 mo sample
# two 24 mo samples are duplicates I am dropping the duplicate - should it be completely removed? 
ACNC_24mo =
  ACNC_24mo %>%
  distinct(Sample_Name, .keep_all = TRUE) %>% # it keeps only the first occurrence
  filter(!str_extract(Sample_Name, "^[^ ]+") %in% remove)

```

## Then apply the LOD cutoff that we've been using - 0.1ng/g - convert values below this to missing
instead will be using corresponding LOQ values 
##Then create a table that presents the rate of detection >= the LOD for each PFAS, separately for the 6mo ad 24mo samples
##Then impute those the missing values with 0.01/sqrt(2)

```{r}
# first changing the NR values to "0" won't let me change it directly to NA 
ACNC_6mo <- ACNC_6mo %>%
  mutate(across(everything(), ~ ifelse(. == "NR", "0", .))) %>% 
  mutate(across(-1, as.numeric))  # setting all columns to numeric except first

  
ACNC_24mo <- ACNC_24mo %>%
  mutate(across(everything(), ~ ifelse(. == "NR", "0", .))) %>%
  mutate(across(-1, as.numeric))  # setting all columns to numeric except first



#summing values also and saving them in dataframe above_LOD
above_LOD <- data.frame(
   mo6 = numeric(),
   mo24 = numeric(),
  stringsAsFactors = FALSE
)

#using a for loop to make values less than approprite threshold NA 
for( p in colnames(ACNC_6mo)[2:36]){
  LOQ = thresholds[p, 2]
  imputed_value = LOQ/sqrt(2)
  
  ACNC_6mo[,p] = if_else(ACNC_6mo[,p] < LOQ , NA, ACNC_6mo[,p])
  ACNC_24mo[,p] = if_else(ACNC_24mo[,p] < LOQ , NA, ACNC_24mo[,p])
  
  above_LOD[p,"mo6"] <- sum(ACNC_6mo[,p] >= LOQ, na.rm = TRUE)
  above_LOD[p,"mo24"] <- sum(ACNC_24mo[,p] >= LOQ, na.rm = TRUE)
  
  # imputation done at the end 
 ACNC_6mo[,p] = replace(ACNC_6mo[,p], is.na(ACNC_6mo[,p]), imputed_value)
 ACNC_24mo[,p] = replace(ACNC_24mo[,p], is.na(ACNC_24mo[,p]), imputed_value) 
}

above_LOD

```




## Then create a large table that presents the geometric mean, geometric standard deviation, and range of absolute values (min:max) for each PFAS, separately for 6mo and 24mo
```{r}
multiple_stats <- function(x) {
      c(geom_mean = exp(mean(log(x))), geom_sd = exp(sd(log(x))), min_val = min(x), max_val = max(x) )}

stats_6mo = t(sapply(ACNC_6mo[,2:36], multiple_stats)) 
stats_6mo
stats_24mo = t(sapply(ACNC_24mo[,2:36], multiple_stats)) 
stats_24mo

```

## Then create some visualization of the PFAS levels overall, and how they change within a participant between 6om and 24mo. Do levels tend to stay stable, or do they tend to increase/decrease? Or is it pretty different for everyone?

```{r}
#ACNC_24mo$length = rep("24mo", 124)
#ACNC_6mo$length = rep("6mo", 124)

datas = bind_rows(ACNC_6mo, ACNC_24mo) %>%
      pivot_longer(
        cols = -starts_with("Sa"), # Select columns to pivot
        names_to = "Chemical",         # New column for original column names
        values_to = "Concentration"          # New column for values
      ) %>% 
     mutate(Time_Point = if_else(str_detect(Sample_Name, "24mo"), "24mo", "6mo"))  %>% 
  mutate(Chemical_Class = case_when(
    str_detect(Chemical, "PFOA|PFTr|PFTe|PFPe|PFNA|PFDo|PFUd|PFDA|PFBA|PFHxA|PFHpA") ~ "PFCA",
    str_detect(Chemical, "PFOS|PFHxS|PFDS|PFBS|PFNS|PFHpS") ~ "PFSA",
    str_detect(Chemical, "FTS") ~ "FTS",
    str_detect(Chemical, "FOSA") ~ "FOSAs",
    TRUE ~ "Other"
  ))

datas$Sample_Name = str_extract(datas$Sample_Name, "^[^ ]+")
    
```


```{r}
ggplot(datas, aes(x = Chemical, y = log10(Concentration), fill = Time_Point)) +
  geom_boxplot(outlier.shape = NA) +
  coord_flip() +
  facet_wrap(~ Chemical_Class, scales = "free_y") +
 scale_fill_brewer(palette = "Set1") +
  labs(
    x = "PFAS",
    y = expression(Log[10]~Concentration),
    fill = "Time Point",
    title = "PFAS Concentrations at 6 months vs 24 months"
  )  +
  theme_bw(base_size = 14) +
  theme(
    axis.text.y = element_text(size = 8),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5)
  )

  
```

```{r}
datas$Time_Point <- factor(datas$Time_Point, levels = c("6mo", "24mo"))
avg_diff <- datas %>%
  group_by(Chemical, Sample_Name) %>%
  summarise(diff = diff(log10(Concentration[order(Time_Point)])), .groups = "drop") %>%
  group_by(Chemical) %>%
  summarise(mean_diff = mean(diff, na.rm = TRUE))
```


```{r}
pdf("avg_diff_pfas.pdf", width = 8, height = 6) 

facet_levels <- unique(datas$Chemical)

for (level in facet_levels) {
  subset_data <- subset(datas, Chemical == level)

  # mean difference for this chemical
  diff_val <- avg_diff$mean_diff[avg_diff$Chemical == level]
  
  p <- ggplot(subset_data, aes(x=Time_Point, y=log10(Concentration), group=Sample_Name, colour=Sample_Name)) +
    geom_point(size = 2, alpha = 0.7) + 
    geom_line(alpha = 0.5) +
    theme_bw(base_size = 14) +
    theme(legend.position = "none") +
    labs(
      title = paste("PFAS:", level),
      subtitle = paste("Mean Δ (6mo - 24mo):", round(diff_val, 2)),
      x = "Time Point", 
      y = expression(Log[10]~Concentration)
    )
  
  print(p)
}
dev.off()
```



```{r}
above_LOD$percent_mo24 = above_LOD$mo24/124*100
above_LOD$percent_mo6 = above_LOD$mo6/124*100

pfas_above70_mo6 = rownames(above_LOD)[which(above_LOD$percent_mo6 > 70)]

pfas_above70_mo24 =rownames(above_LOD)[which(above_LOD$percent_mo24 > 70)]

# both mo6 and mo24 have the same pfas above 70%


```
# would be pairwise correlations for each PFAS: scatter plots with 6mo on the x axis and 24mo on the y-axis, then correlations coefficient and p-values. Only do this for the PFAS that are well detected (>70%).


```{r}
pfas_cor = cor(ACNC_6mo[,pfas_above70_mo6], ACNC_24mo[,pfas_above70_mo24])
```



```{r}
mo24 = ACNC_24mo %>% mutate(Sample_Name = str_extract(Sample_Name, "^[^ ]+")) %>% arrange(Sample_Name)  %>% select( contains("Sample") | contains(pfas_above70_mo24) ) 
colnames(mo24)[2:8] = sapply(colnames(mo24)[2:8], function(name) paste0("mo24_", name))

subset = ACNC_6mo %>% select( contains("Sample") | contains(pfas_above70_mo24) ) %>% left_join(mo24, by = "Sample_Name") %>% mutate(across(-1, ~ log(.)))


```


```{r}
pdf(file = "pairwise_correlations.pdf") 

for( i in pfas_above70_mo24){
 p =  ggplot(subset, aes_string(x = i, y = paste0("mo24_",i))) +
      geom_point() + # Add scatter points
      geom_smooth(method = "lm", se = FALSE, color = "blue") + # Add linear regression line without standard error
      stat_cor(method = "spearman", label.x.npc = "left", label.y.npc = "top")  +
      labs(x = "6 months", y = "24 months", title = paste("Correlation between 6 month and 24 month", i))
 
 print(p) 
}
dev.off()

```



